---
title: "data_analysis"
output: html_document
date: "2024-03-12"
---

# What can be said about the quality of the data?

Let's do some sanity checks. First, we'd like to look at the distributions of all individual channels, and compare these between replicate.

This will enable us to get a sense of what the phenotypes of these cells look like, in terms of central tendencies, and variability.

Also, that will enable us to control for reproducibility.

# 1D Distributions

First, let's write a function that plots all desired distributions.
```{r message=FALSE, warning=FALSE}
plot_all_1D_distributions <- function(ini,end,condition){
  cyto_plot_save(sprintf("./figures/1d_distributions_%s.pdf",condition),width=25,height=25)
  cyto_plot_custom(layout=c(6,5))
  
  xlim_inf <- rep(c(0),times=9)
  xlim_sup <- c(4e6,4e6,1e6,1e6,1e6,7e5,2e6,7e5,1e6)
  
  for (i in 1:9){
    cyto_plot(fs[ini:end],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title=condition,xlim=c(xlim_inf[i],xlim_sup[i])
              )}
  for (i in 10:30){
    cyto_plot(fs[ini:end],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title=condition,xlim=c(-1e5,1e6))}
  cyto_plot_complete()}
```

Then we plot all 1D distributions for all "postSpn","preSpn" and "saline" replicates.

```{r}
plot_all_1D_distributions(1,6,"postSpn")
plot_all_1D_distributions(7,12,"preSpn")
plot_all_1D_distributions(13,18,"saline")
```

Overall, postSpn samples are more reproducible than preSpn and saline. That also can be explained by the actual number of cells there are in each sample. There are 10 times more cells in postSpn versus preSpn, and 20 times more cells in postSpn versus saline.

```{r}
cyto_stats_compute(fs,alias="root",stat="count")
cyto_stats_compute(fs,alias="root",stat="count") %>% 
  group_by(treatment) %>% 
  summarise(n=sum(Count))
```

Let's compare these distributions in each channel: pre|saline|post, to determine where distributions have changed.

First we build the desired function.

```{r message=FALSE, warning=FALSE}
plot_1D_distribution_comparison <- function(){
  cyto_plot_save("./figures/1d_distribution_comparison.pdf",width=15,height=45*3)
  cyto_plot_custom(layout=c(30,3))
  
  xlim_inf <- rep(c(0),times=9)
  xlim_sup <- c(4e6,4e6,1e6,1e6,1e6,7e5,2e6,7e5,1e6)
  
  for (i in 1:9){
    cyto_plot(fs[7:12],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title="preSpn",xlim=c(xlim_inf[i],xlim_sup[i]))
    cyto_plot(fs[13:18],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title="saline",xlim=c(xlim_inf[i],xlim_sup[i]))
    cyto_plot(fs[1:6],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title="postSpn",xlim=c(xlim_inf[i],xlim_sup[i]))
    }
  for (i in 10:30){
    cyto_plot(fs[7:12],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,
              xlim=c(-1e5,1e6),title="preSpn")
    cyto_plot(fs[13:18],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,
              xlim=c(-1e5,1e6),title="saline")
    cyto_plot(fs[1:6],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,
              xlim=c(-1e5,1e6),title="postSpn")}
  cyto_plot_complete()}
```

Then we plot all possible comparisons.

```{r}
plot_1D_distribution_comparison()
```

Let's plot here only interesting examples.
First we write a dedicated function.

```{r}
plot_1D_distribution_comparison_by_name <- function(channel_list){
  #cyto_plot_save("./figures/1d_distribution_comparison_.pdf",width=15,height=45*3)
  
  xlim_inf <- c(rep(c(0),times=9),rep(c(-1e5),times=20))
  xlim_sup <- c(4e6,4e6,1e6,1e6,1e6,7e5,2e6,7e5,1e6,rep(c(1e6),times=20))
  markers <- simplify(lapply(c(1:length(cyto_markers(fs))), FUN=function(.i) return(cyto_markers(fs)[[.i]])))
  factor_list <- factor(channel_list,levels=channel_list)
  indexes <- which(simplify(base::lapply(markers,FUN=function(.x) .x %in% channel_list)))
  markers <- factor(markers,levels=channel_list)
  
  indexes <- tibble('markers'=markers,'indexes'=c(1:length(markers))) %>% 
    arrange(markers) %>% 
    na.omit() %>% 
    .$indexes
  
  for(i in indexes){
    cyto_plot_custom(layout=c(1,3))
    cyto_plot(fs[7:12],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title="preSpn",xlim=c(xlim_inf[i],xlim_sup[i]))
    cyto_plot(fs[13:18],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title="saline",xlim=c(xlim_inf[i],xlim_sup[i]))
    cyto_plot(fs[1:6],parent="root",channels=cyto_channels(fs)[i],density_stack=0.7,axes_trans=transList,title="postSpn",xlim=c(xlim_inf[i],xlim_sup[i]))
    cyto_plot_complete()}}
```

Then the plotting.

```{r,warning=FALSE}
plot_1D_distribution_comparison_by_name(c("CD69"))
plot_1D_distribution_comparison_by_name(c("CD44"))
plot_1D_distribution_comparison_by_name(c("PDL2"))
```

# 2D distributions

Let's make one PDF per file, where we look at all 2D distributions.

```{r message=FALSE, warning=FALSE}
plot_2d_distributions <- function(k){
  
  condition <- cyto_details(fs)$conditions[k]
  cyto_plot_save(sprintf("./figures/2d_heatmap_%s.pdf",condition),width=135,height=135)
  cyto_plot_custom(layout=c(30,30))
  for(i in 1:30){
    for(j in 1:30){
      cyto_plot(fs[k],parent="root",channels=cyto_channels(fs)[c(i,j)],
                xlim=c(lim_inf[i],lim_sup[i]),
                ylim=c(lim_inf[j],lim_sup[j]),
                title=cyto_details(fs)$conditions[k],
                axes_trans=transList)
    }
  }
  cyto_plot_complete()
}

for(p in 1:18){
  plot_2d_distributions(p)}
```

Here, let's write some code to plot single 2D distributions.

```{r}
plot_single_2d_distributions <- function(k,x_marker,y_marker){
  
  condition <- cyto_details(fs)$conditions[k]
  #cyto_plot_save(sprintf("./figures/2d_heatmap_%s_%s_%s.pdf",x_marker,y_marker,condition),width=135,height=135)
  cyto_plot_custom(layout=c(1,1))
  
  markers <- simplify(lapply(c(1:length(cyto_markers(fs))), FUN=function(.i) return(cyto_markers(fs)[[.i]])))
  i <- which(simplify(base::lapply(markers,FUN=function(.x) .x %in% x_marker)))
  j <- which(simplify(base::lapply(markers,FUN=function(.y) .y %in% y_marker)))

  cyto_plot(fs[k],parent="root",channels=cyto_channels(fs)[c(i,j)],
            xlim=c(lim_inf[i],lim_sup[i]),
            ylim=c(lim_inf[j],lim_sup[j]),
            title=cyto_details(fs)$conditions[k],
            axes_trans=transList)
  cyto_plot_complete()
}
```

Plot single 2D distributions.

```{r,warning=FALSE}
plot_single_2d_distributions(1,"CD69","CD44")
```

# Grouped 2D distributions for comparison

```{r message=FALSE, warning=FALSE}
plot_2d_distributions_comparison <- function(n){
  
  lim_inf <- c(rep(c(0),times=9),rep(-1e5,times=21))
  lim_sup <- c(4e6,4e6,1e6,1e6,1e6,7e5,2e6,7e5,1e6,rep(1e6,times=21))
  
  cyto_plot_save(sprintf("./figures/2d_comparison_%s.pdf",cyto_markers(fs)[n]),width=10,height=100)
  cyto_plot_custom(layout=c(30,3))
  #for(i in 1:30){
    i <- n
    for(j in 1:30){
      cyto_plot(fs[7:12],parent="root",channels=cyto_channels(fs)[c(i,j)],
                xlim=c(lim_inf[i],lim_sup[i]),
                ylim=c(lim_inf[j],lim_sup[j]),
                axes_trans=transList,
                group_by = c("treatment"),
                point_col_alpha = 0.5,
                #point_col="blue",
                contour_lines = 10,
                contour_line_type = 1,
                contour_line_width = 1.2,
                contour_line_col = "black",
                contour_line_alpha = 0.2,
                display=1500,
                title=c("preSpn"))
      cyto_plot(fs[13:18],parent="root",channels=cyto_channels(fs)[c(i,j)],
                xlim=c(lim_inf[i],lim_sup[i]),
                ylim=c(lim_inf[j],lim_sup[j]),
                axes_trans=transList,
                group_by = c("treatment"),
                point_col_alpha = 0.5,
                #point_col="blue",
                contour_lines = 10,
                contour_line_type = 1,
                contour_line_width = 1.2,
                contour_line_col = "black",
                contour_line_alpha = 0.2,
                display=1500,
                title=c("saline"))
      cyto_plot(fs[1:6],parent="root",channels=cyto_channels(fs)[c(i,j)],
                xlim=c(lim_inf[i],lim_sup[i]),
                ylim=c(lim_inf[j],lim_sup[j]),
                axes_trans=transList,
                group_by = c("treatment"),
                point_col_alpha = 0.5,
                #point_col="blue",
                contour_lines = 10,
                contour_line_type = 1,
                contour_line_width = 1.2,
                contour_line_col = "black",
                contour_line_alpha = 0.2,
                display=1500,
                title=c("postSpn"))
    }
  cyto_plot_complete()
}

#plot_2d_distributions_comparison(1)
```

```{r}
for(n in 1:30){
  plot_2d_distributions_comparison(n)}
```

# Check gating schemes

```{r message=FALSE, warning=FALSE}
cyto_plot_gating_scheme(fs,
                        #channels=c("CD19","7AAD"),
                        #alias=c("B-cells","T-cells"),
                        xlim=c(-1e5,1e6),
                        ylim=c(-1e5,1e6))
```
  
```{r message=FALSE, warning=FALSE}
cyto_plot_gating_scheme(fs[[1]],
                        back_gate = TRUE,
                        gate_track = TRUE)
```

```{r}
count_df <- cyto_stats_compute(fs,
                   alias = c("IgD_CD8+IgM+","IgD_CD8-IgM+","IgD_CD8-IgM-","CD8_IgD+","CD4+","CD4-CD8-"),
                   stat = "count")

freq_df <- cyto_stats_compute(fs,parent="root",
                   alias = c("IgD_CD8+IgM+","IgD_CD8-IgM+","IgD_CD8-IgM-","CD8_IgD+","CD4+","CD4-CD8-"),
                   stat = "freq")
```
    
```{r}
freq_df %>% 
  dplyr::filter(replica %in% as.character(c(79:84))) %>% 
  ggplot(aes(x=Population,y=Frequency,fill=replica))+
  geom_col(position = "dodge")+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="postSpn")

freq_df %>% 
  dplyr::filter(replica %in% as.character(c(1:6))) %>% 
  ggplot(aes(x=Population,y=Frequency,fill=replica))+
  geom_col(position = "dodge")+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="preSpn")

freq_df %>% 
  dplyr::filter(replica %in% as.character(c(67:72))) %>% 
  ggplot(aes(x=Population,y=Frequency,fill=replica))+
  geom_col(position = "dodge")+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="saline")
```
```{r}
freq_df %>% 
  #dplyr::filter(replica %in% as.character(c(79:84))) %>% 
  group_by(treatment,Population) %>%
  mutate(m_freq=mean(Frequency,na.rm=TRUE),
         sd_freq=sd(Frequency,na.rm=TRUE)) %>% 
  ungroup() %>% 
  distinct(treatment,Population,.keep_all = TRUE) %>% 
  ggplot(aes(x=Population,y=m_freq,fill=treatment))+
  geom_col(position = "dodge")+
  geom_errorbar(aes(ymin=m_freq-sd_freq,ymax=m_freq+sd_freq),position='dodge')+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Population Mean frequency")+
  labs(fill="Condition")
ggsave("./figures/mean_frequencies.pdf",height=5,width=7)
```
    
    
    
```{r}
count_df %>% 
  dplyr::filter(replica %in% as.character(c(79:84))) %>% 
  ggplot(aes(x=Population,y=Count,fill=replica))+
  geom_col(position = "dodge")+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="postSpn")+
  scale_y_log10()

count_df %>% 
  dplyr::filter(replica %in% as.character(c(1:6))) %>% 
  ggplot(aes(x=Population,y=Count,fill=replica))+
  geom_col(position = "dodge")+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="preSpn")+
  scale_y_log10()

count_df %>% 
  dplyr::filter(replica %in% as.character(c(67:72))) %>% 
  ggplot(aes(x=Population,y=Count,fill=replica))+
  geom_col(position = "dodge")+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title="saline")+
  scale_y_log10()
```

```{r}
count_df %>% 
  #dplyr::filter(replica %in% as.character(c(79:84))) %>% 
  group_by(treatment,Population) %>%
  mutate(m_count=mean(Count,na.rm=TRUE),
         sd_count=sd(Count,na.rm=TRUE)) %>% 
  ungroup() %>% 
  distinct(treatment,Population,.keep_all = TRUE) %>% 
  ggplot(aes(x=Population,y=m_count,fill=treatment))+
  geom_col(position = "dodge")+
  geom_errorbar(aes(ymin=m_count-sd_count,ymax=m_count+sd_count),position='dodge')+
  #facet_wrap(~replica)+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Population Mean Count")+
  labs(fill="Condition")+
  scale_y_log10()
ggsave("./figures/mean_counts.pdf",height=5,width=7)
```
